---
output:
  pdf_document: default
  html_document: default
---
```{r}
load("synthetic_raw.Rda")
library(SmoothHOOI)
library(ggplot2)
library(dplyr)
library(tidyr)
```

This is an illustration of how to use our functions on synthetic data 

```{r}
set.seed(32123)
```

# Generate synthetic data
```{r}
#synthetic <- synthetic_data(L_tilde, R_tilde, mean_G, cov_G, E, p=207, noise_level=1, pattern="random", percent=0.2)

synthetic <- synthetic_data(L_tilde, R_tilde, mean_G, cov_G, E, p=207, noise_level=1, pattern = "structured")

synthetic_dat <- synthetic$sim_Mmiss

#save(synthetic_dat, file = "synthetic_data.Rda")
```

# Make the second order difference matrix 
```{r}
D2 <- SecDiffMat(24)
```

# Coarse-to-fine grid search (Example)
```{r}
original <- (synthetic_dat@data[ , 1, 180])*12+70
lambda_seq <- c(0.1, 1, 10, 50, 100, 500)
est_df <- matrix(NA, nrow=24, ncol=length(lambda_seq))
for (i in 1:length(lambda_seq)){
  lambda <- lambda_seq[i]
  res <-  mglram(tnsr = synthetic_dat@data, ranks = c(6, 3), init=0, D = D2,
                 lambda = lambda, max_iter = 500, tol = 1e-5, L0_ = NULL)
  est_df[,i] <- (res$est[,1,180])*12 +70
}
org_df <- data.frame(cbind(c(1:24), original))
est_df <- data.frame(cbind(c(1:24), est_df)) 
colnames(org_df) <- c("hour", "original")
colnames(est_df) <- c("hour", paste0("lambda",lambda_seq))
est_long <- pivot_longer(est_df, cols = starts_with("lambda"),
                         names_to = "lambda", values_to = "y")
est_long$lambda <- factor(est_long$lambda, levels=c("lambda0.1","lambda1","lambda10","lambda50","lambda100", "lambda500"))
```

```{r, warning=FALSE}
custom_colors <- c("grey","#9ECAE1","#6BAED6", "#4292C6", "#2171B5","darkblue")
names(custom_colors) <- levels(factor(est_long$lambda))  

ggplot() +
  geom_point(data = org_df, aes(x = hour, y = original), color = "black") +  # Original data
  geom_line(data = est_long, aes(x = hour, y =y, color = lambda)) +  # Estimated curves
  labs(title = "",
       x = "HOUR", y = "DBP") +
  scale_color_manual(values = custom_colors,
                    name = expression(lambda)) + 
  theme(legend.position = c(1, 1),        
    legend.justification = c(1, 1), 
    panel.grid = element_blank(),             
    panel.background = element_blank(),       
    plot.background = element_blank(),        
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),  # add border rectangle
    axis.line = element_blank(),
    legend.background = element_blank(),  # removes the outer background & border
    legend.key = element_blank(),
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 16),
    legend.text = element_text(size = 16,hjust=0),
    legend.title = element_text(size = 16)
    )

```

```{r}
original <- (synthetic_dat@data[ , 1, 1])*12 + 70
r1_seq <- c(3,6,9,12)
est_df <- matrix(NA, nrow=24, ncol=length(r1_seq))

for (i in 1:length(r1_seq)){
  r1 <- r1_seq[i]
  res <-  mglram(tnsr = synthetic_dat@data, ranks = c(r1, 3), init=0, D = D2,
                 lambda = 10, max_iter = 500, tol = 1e-5, L0_ = NULL)
  est_df[,i] <- (res$est[,1,1])*12 + 70
}

colnames(est_df) <- paste0("r", r1_seq)
est_df <- cbind(hour = 1:24, est_df)
est_df <- data.frame(est_df)
org_df <- data.frame(hour = 1:24, original = original)
est_long <- pivot_longer(est_df, cols = starts_with("r"),
                         names_to = "r", values_to = "y")
est_long$r <- factor(est_long$r, levels = paste0("r", r1_seq))
```

```{r, warning=FALSE}
custom_colors <- c("#FEC44F", "#FE9929","#EC7014","#CC4C02")
names(custom_colors) <- levels(factor(est_long$r))  

ggplot() +
  geom_point(data = org_df, aes(x = hour, y = original), color = "black") +  # Original data
  geom_line(data = est_long, aes(x = hour, y =y, color = r)) +  # Estimated curves
  labs(title = "",
       x = "HOUR", y = "DBP") +
  scale_color_manual(values = custom_colors,
                     labels = c("r3" = 3, "r6" = 6,
                                "r9" = 9, "r12" = 12),
                    name =  expression(r[1])) + 
  theme(legend.position = c(1, 1),        
    legend.justification = c(1, 1), 
    panel.grid = element_blank(),             
    panel.background = element_blank(),       
    plot.background = element_blank(),        
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),  # add border rectangle
    axis.line = element_blank(),
    legend.background = element_blank(),  # removes the outer background & border
    legend.key = element_blank(),
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 16),
    legend.text = element_text(size = 16,hjust=0),
    legend.title = element_text(size = 16)
    )

```

# Find optimal hyperparameter using k-fold cross-validation
```{r}
start_time <- Sys.time()

kcv_res <- kcv(tnsr=synthetic_dat@data, rank_grid=as.matrix(expand.grid(r1<-seq(3,6,by=1), r2<-c(2,3))), lambda_seq=seq(1,30,by=1),
               k=5, L0_ =NULL, D=D2, tol=0.01, max_iter=500, init=0)

end_time <- Sys.time()

exec_time <- end_time - start_time
print(exec_time)

kcv_res
```

# Check explained variability

```{r}
opt_res <- mglram(tnsr = synthetic_dat@data, ranks = c(4, 2), init=0, D = D2,
       lambda = 22, max_iter = 500, tol = 1e-5, L0 = NULL)
```

```{r}
# Cumulative explained variability when r1 increases
tilde <- MakeIdent(opt_res$L, opt_res$G, opt_res$R)
L_tilde <- tilde$L_tilde
G_tilde <- tilde$G_tilde
R_tilde <- tilde$R_tilde

nmiss_idx <- which(!is.na(synthetic_dat@data))

cum_var_L <- rep(NA, 4)
for (i in 1:4){
  comp <- array(NA, c(24,3,207))
  for (j in 1:207){
    comp[ , , j] <- matrix(L_tilde[, 1:i], ncol=i) %*% matrix(G_tilde[1:i, , j], nrow=i) %*% t(R_tilde)
  }
  cum_var_L[i] <- sum(comp[nmiss_idx]^2)/sum(synthetic_dat@data[nmiss_idx]^2)
}

cum_var_L # 0.4472346 0.5592966 0.6085780 0.6106854
```

```{r}
# Separate explained variability for different L ranks
cum_var_L0 <- c(0, cum_var_L[1:3])
sep_var_L <- cum_var_L - cum_var_L0

sep_var_L # 0.447234584 0.112062011 0.049281371 0.002107439
```

```{r}
# Cumulative explained variability when r2 increases
cum_var_R <- rep(NA, 2)
for (i in 1:2){
  comp <- array(NA, dim(synthetic_dat@data))
  for (j in 1:207){
    comp[ , , j] <- L_tilde %*% matrix(G_tilde[,1:i, j], ncol=i) %*% matrix(t(R_tilde[, 1:i]), nrow=i)
  }
  cum_var_R[i] <- sum(comp[nmiss_idx]^2)/sum(synthetic_dat@data[nmiss_idx]^2)
}

cum_var_R # 0.3812621 0.6106854
```

```{r}
# Separate explained variability for different R ranks
cum_var_R0 <- c(0, cum_var_R[1:1])
sep_var_R <- cum_var_R - cum_var_R0

sep_var_R # 0.3812621 0.2294233
```

```{r}
# Plot the explained variability
var_data_L <- data.frame(
  component = factor(seq_along(sep_var_L), levels = seq_along(sep_var_L)), 
  value = sort(sep_var_L, decreasing = TRUE) 
)

var_data_L$percentage <- paste0(round(var_data_L$value * 100, 2), "%")

ggplot(var_data_L, aes(x = component, y = value)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = percentage), vjust = -0.5, size = 5) +
  geom_line(aes(group = 1), color = "black", linewidth = 0.5) +
  geom_point(color = "black", size = 1) +
  labs(y = "Explained Variability", x="L Component", title="") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title.y = element_text(size = 15),
        axis.title.x = element_text(size = 15), 
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) 


```

```{r}
# Plot the explained variability
var_data_R <- data.frame(
  component = factor(seq_along(sep_var_R), levels = seq_along(sep_var_R)), 
  value = sort(sep_var_R, decreasing = TRUE) 
)

var_data_R$percentage <- paste0(round(var_data_R$value * 100, 2), "%")

ggplot(var_data_R, aes(x = component, y = value)) +
  geom_bar(stat = "identity", fill = "skyblue", width = 0.4) +
  geom_text(aes(label = percentage), vjust = -0.5, size = 5) +
  geom_line(aes(group = 1), color = "black", linewidth = 0.5) +
  geom_point(color = "black", size = 1) +
  labs(title = "", x = "R Component", y = "Explained Variability") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.y = element_text(size = 15),
        axis.title.x = element_text(size = 15), 
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

# Update hyperparameters for parsimony
```{r}
start_time <- Sys.time()

res <- mglram(tnsr = synthetic_dat@data, ranks = c(3, 2), init=0, D = D2,
       lambda = 24, max_iter = 500, tol = 1e-5, L0 = NULL)

res$conv

end_time <- Sys.time()

exec_time <- end_time - start_time
print(exec_time)


tilde <- MakeIdent(L=res$L, G=res$G, R=res$R)

L_tilde <- tilde$L_tilde
R_tilde <- tilde$R_tilde
G_tilde <- tilde$G_tilde

L_tilde
R_tilde 
G_tilde[ , ,1:5]

```

```{r}
dataL = data.frame(loading = c(L_tilde[, 1], L_tilde[, 2], L_tilde[, 3]),
                   component = c(rep("1st", 24), rep("2nd", 24), rep("3rd", 24)),
                   hour = rep(12:35, 3))

pL = dataL %>%
  ggplot(aes(x = hour, y = loading)) + geom_line(alpha = 0.5) + geom_point() + geom_hline(yintercept = 0, color = "red") +
  scale_x_continuous(breaks = c(12, 15, 18, 21, 24, 27, 30, 33, 36), labels = c(12, 15, 18, 21, 0, 3, 6, 9, 12)) +
  xlab("Hour of the day") + 
  facet_grid(~ component) + 
  ylab("Coefficient") +
  theme(text = element_text(size = 15)) +
  scale_y_continuous(limits = c(-0.4, 0.4))

pL
```

```{r}
R_tilde[, 1] # first component of R tilde
R_tilde[, 2] # second component of R tilde

## 36/100 DBP + 31/50 SBP + 33/100 HR (Seem to be joint effect of the three)
R_tilde[1, 1]/sum(R_tilde[, 1]) # DBP: 36/100 
R_tilde[2, 1]/sum(R_tilde[, 1]) # SBP: 31/100 
R_tilde[3, 1]/sum(R_tilde[, 1]) # HR: 33/100 

## 17/100 DBP + 33/100 SBP - 50/100 HR (Contrast between blood pressure and heart rate)
R_tilde[1, 2]/sum(abs(R_tilde[, 2])) # DBP: 17/100
R_tilde[2, 2]/sum(abs(R_tilde[, 2])) # SBP: 33/100 
R_tilde[3, 2]/sum(abs(R_tilde[, 2])) # HR: -50/100 
```

```{r}
# Prepare to plot effect of time components 
n = 207

Ri = array(NA, dim = c(3, 3, n))
for (i in 1:n){
  Ri[, , i] = G_tilde[, , i] %*% t(R_tilde)
}

RiL1 = Ri[1, , ] 
apply(RiL1, 1, sd) # 2.789581 2.997533 3.661661 (sd for DBP, SBP, and HR for 1st time component)

RiL2 = Ri[2, , ]
apply(RiL2, 1, sd) # 1.379921 1.401255 1.444989 (sd for DBP, SBP, and HR for 2nd time component)

RiL3 = Ri[3, , ]
apply(RiL3, 1, sd) # 1.067016 1.115904 1.072485 (sd for DBP, SBP, and HR for 3rd time component)


DBPmean = rep(70, 24) + (L_tilde[, 1] * mean(RiL1[1,]) + L_tilde[, 2] * mean(RiL2[1,]) + L_tilde[, 3] * mean(RiL3[1,])) * 12 # 70 is the mean of DBP, 12 is the sd of DBP, doing this because we normalized the data initially
SBPmean = rep(129, 24) + (L_tilde[, 1] * mean(RiL1[2,]) + L_tilde[, 2] * mean(RiL2[2,]) + L_tilde[, 3] * mean(RiL3[2,])) * 18 # 129 is the mean of SBP, 18 is the sd of SBP
HRmean = rep(76, 24) + (L_tilde[, 1] * mean(RiL1[3,]) + L_tilde[, 2] * mean(RiL2[3,]) + L_tilde[, 3] * mean(RiL3[3,])) * 13 # 76 is the mean of HR, 13 is the sd of HR 
```

```{r}
# Plot effect of time components on DBP 
dataLeffects = data.frame(DBPmean = rep(DBPmean, 6), 
                          value = c(DBPmean + L_tilde[, 1] * 3*12, DBPmean - L_tilde[, 1] * 3*12, 
                                    DBPmean + L_tilde[, 2] *1.1*12, DBPmean - L_tilde[, 2] * 1.1*12, 
                                    DBPmean + L_tilde[, 3] * 0.9*12, DBPmean - L_tilde[, 3] * 0.9*12),
                          component = c(rep("1st", 24 * 2), rep("2nd", 24 * 2), rep("3rd", 24 * 2)),
                          sign = rep(rep(c("positive", "negative"), each = 24), 3),
                          hour = rep(12:35, 6))

pdataL = dataLeffects %>%
  ggplot(aes(x = hour, y = value, group = sign, col = sign)) + geom_line(alpha = 0.5) +
  scale_x_continuous(breaks = c(12, 15, 18, 21, 24, 27, 30, 33, 36), labels = c(12, 15, 18, 21, 0, 3, 6, 9, 12)) +
  xlab("Hour of the day") + 
  facet_grid(~ component) + 
  ylab("DBP") + geom_line(aes(x = hour, y = DBPmean), col = "black") +
  theme(text = element_text(size = 15)) 

pdataL
```

```{r}
# Plot effect of time components on SBP 
dataLeffects = data.frame(SBPmean = rep(SBPmean, 6), 
                          value = c(SBPmean + L_tilde[, 1] * 3.1*18, SBPmean - L_tilde[, 1] * 3.1*18, 
                                    SBPmean + L_tilde[, 2] * 1.2*18, SBPmean - L_tilde[, 2] * 1.2*18, 
                                    SBPmean + L_tilde[, 3] * 1*18, SBPmean - L_tilde[, 3] * 1*18),
                          component = c(rep("1st", 24 * 2), rep("2nd", 24 * 2), rep("3rd", 24 * 2)),
                          sign = rep(rep(c("positive", "negative"), each = 24), 3),
                          hour = rep(12:35, 6))

pdataL = dataLeffects %>%
  ggplot(aes(x = hour, y = value, group = sign, col = sign)) + geom_line(alpha = 0.5) +
  scale_x_continuous(breaks = c(12, 15, 18, 21, 24, 27, 30, 33, 36), labels = c(12, 15, 18, 21, 0, 3, 6, 9, 12)) +
  xlab("Hour of the day") + facet_grid(~ component) + ylab("SBP") + geom_line(aes(x = hour, y = SBPmean), col = "black") +
  theme(text = element_text(size = 15)) 

pdataL
```

```{r}
# Plot effect of time components on HR
dataLeffects = data.frame(HRmean = rep(HRmean, 6), 
                          value = c(HRmean + L_tilde[, 1] * 4*13, HRmean - L_tilde[, 1] *4*13, 
                                    HRmean + L_tilde[, 2] * 1.3*13, HRmean - L_tilde[, 2] * 1.3*13, 
                                    HRmean + L_tilde[, 3] * 1.1*13, HRmean - L_tilde[, 3] * 1.1*13),
                          component = c(rep("1st", 24 * 2), rep("2nd", 24 * 2), rep("3rd", 24 * 2)),
                          sign = rep(rep(c("positive", "negative"), each = 24), 3),
                          hour = rep(12:35, 6))

pdataL = dataLeffects %>%
  ggplot(aes(x = hour, y = value, group = sign, col = sign)) + geom_line(alpha = 0.5) +
  scale_x_continuous(breaks = c(12, 15, 18, 21, 24, 27, 30, 33, 36), labels = c(12, 15, 18, 21, 0, 3, 6, 9, 12)) +
  xlab("Hour of the day") + facet_grid(~ component) + ylab("HR") + geom_line(aes(x = hour, y = HRmean), col = "black") +
  theme(text = element_text(size = 15)) 

pdataL
```


# Plot estimated curves for some patients
```{r}
for (i in 1:5){
  plot(synthetic_dat@data[ , 1, i], main=paste(i,"th patient"), xlab="Hour", ylab="DBP", ylim=c(-2,2)) # original data
  lines(res$est[ , 1, i], col="blue") # estimated curve by our algorithm
}
```

```{r}
# Plot Estimated Curves for SBP
for (i in 70:75){
  plot(synthetic_dat@data[ , 2, i], main=paste(i,"th patient"), xlab="Hour", ylab="SBP", ylim=c(-2.5,2))
  lines(res$est[ , 2, i], col="blue")
}
```

```{r}
# Plot Estimated Curves for HR
for (i in 170:175){
  plot(synthetic_dat@data[ , 3, i], main=paste(i,"th patient"), xlab="Hour", ylab="HR")
  lines(res$est[ , 3, i], col="blue")
}
```

# Comparison with FPCA (Example)
```{r, warning=FALSE}
library(refund)
h1 <- rTensor::unfold(synthetic_dat[,1,], row_idx = 2, col_idx = 1)@data
outfpca1 <- refund::fpca.sc(Y = as.matrix(h1), npc = 3, center=FALSE)

res <-  mglram(tnsr = synthetic_dat@data, ranks = c(6, 3), init=0, D = D2,
                 lambda = 12, max_iter = 500, tol = 1e-5, L0 = NULL)

for (i in c(111)){
df <- data.frame(cbind(c(1:24), (res$est[,1,i])*12+70, (outfpca1$Yhat[i,])*12+70))
colnames(df) <- c("hour", "my", "fpca")
original <- (synthetic_dat@data[ , 1, i])*12+70
org_df <- data.frame(cbind(c(1:24), original))
colnames(org_df) <- c("hour", "original")

df_long <- pivot_longer(df, cols=c("my", "fpca"), names_to = "method")

custom_colors <- c("blue","red")
names(custom_colors) <- c("my", "fpca")

p <- ggplot() +
  geom_point(data = org_df, aes(x = hour, y = original), color = "black") +  # Original data
  geom_line(data = df_long, aes(x = hour, y =value, color = method)) +
  labs(title = "",
       x = "HOUR", y = "DBP") +
  scale_color_manual(values = custom_colors,
                     labels = c("my" = "Smooth Tensor Decomposition", "fPCA" = "Functional Principal Component Analysis"),
                    name =  "Method") +
  theme(legend.position = c(1, 1),        
    legend.justification = c(1, 1), 
    panel.grid = element_blank(),             
    panel.background = element_blank(),       
    plot.background = element_blank(),        
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),  # add border rectangle
    axis.line = element_blank(),
    legend.background = element_blank(),  # removes the outer background & border
    legend.key = element_blank(),
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 16),
    legend.text = element_text(size = 14,hjust=0),
    legend.title = element_text(size = 14)
    )
print(p)
}
```

